---
title: "Canonical analysis"
author: "Ehsan Khedive"
date: "2023-10-27"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(tidyverse)
```

## Canonical correspondence analysis

```{r}
library(vegan)
library(vegan3d)
Spider <- read.csv("Makarenkov2002.csv")
Spider %>% dplyr::glimpse()
```


```{r}
lnsp <- log(Spider[,2:13]+1) # log of 0 is inf.
env <- Spider[,c(14:17)]
colnames(env)[1] <- "Water"
env <- as.data.frame(scale(env))
env$Water2 <- env$Water^2
env$Reflection2 <- env$Reflection^2
env$Wateref <- env$Water*env$Reflection
{plot(env$Water,lnsp$Al.fabrilis)
lines(lowess(env$Water,lnsp$Al.fabrilis))}
{plot(env$Reflection,lnsp[,8])
lines(lowess(env$Reflection,lnsp[,8]))}
```

```{r,fig.width=8,fig.height=8}
pcares <- princomp(lnsp, cor = TRUE)
pcares$loadings
pcares$scores
library(factoextra)
fviz_eig(pcares)
biplot(pcares)
fviz_pca_ind(pcares,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
fviz_pca_var(pcares,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```

# 1. Linear Canonical correspondence analysis

```{r}
ccamodel <- cca(lnsp~., env[,1:4])
finalcca <- ordistep(ccamodel, scope=formula(ccamodel))
vif.cca(finalcca)
# vif below 4 indicates no significant relationship between the variables
summary(finalcca)
```

## 1.1. Anova analysis (Permutation test)

```{r}
anova.cca(finalcca)
anova.cca(finalcca, by="terms")
# can remove the variable that is not significant
anova.cca(finalcca, by="axis")
```

## 1.2. Diagnostics tests

```{r}
coef(finalcca)
RsquareAdj(finalcca)$r.squared
RsquareAdj(finalcca)$adj.r.squared
vif.cca(finalcca)
goodness(finalcca,choices = 1:2)
# any species that have coefficient above .6 indicates that the 
# variation in species can be explained by the environmental variables
```

## 1.3. Generating plots

```{r,fig.width=8,fig.height=8}
plot(finalcca, 
scaling = 1, 
display = c("sp", "cn","wa"), 
main = "Triplot CCA Species ~ Environment - scaling 1 - lc scores"
)
# interpretation of plots should be based on the goodness test
```

```{r,fig.width=8,fig.height=8}
plot(finalcca, 
scaling = 2, 
display = c("sp", "lc", "cn"), 
main = "Triplot CCA Species ~ Environment - scaling 2 - lc scores"
)
```

## 1.4. Quadratic model

```{r}
ccamodel2 <- cca(lnsp~., env)
finalcca2<- ordistep(ccamodel2, scope=formula(ccamodel2))
summary(finalcca2)
anova.cca(finalcca2)
anova.cca(finalcca2, by="terms")
anova.cca(finalcca2, by="axis")
coef(finalcca2)
RsquareAdj(finalcca2)$r.squared
RsquareAdj(finalcca2)$adj.r.squared
vif.cca(finalcca2)
goodness(finalcca)
```

```{r,fig.width=8,fig.height=8}
plot(finalcca2, 
scaling = 1, 
display = c("sp", "lc", "cn"), 
main = "Triplot CCA Species ~ Environment - scaling 1 - lc scores"
)
```

```{r,fig.width=8,fig.height=8}
plot(finalcca2, 
scaling = 2, 
display = c("sp", "lc", "cn"), 
main = "Triplot CCA Species ~ Environment - scaling 2 - lc scores"
)
```


Reference:

Borcard, D., Gillet, F., & Legendre, P. (2011). Numerical ecology with R (Vol. 2, p. 688). New York: springer.

Makarenkov, V., & Legendre, P. (2002). Nonlinear redundancy analysis and canonical correspondence analysis based on polynomial regression. Ecology, 83(4), 1146-1161.

Leaw, Chui Pin; Lim, Po Teen; Lee, Li Keat (2020). R script and datasets - CCA. figshare. Dataset. https://doi.org/10.6084/m9.figshare.12356519.v3