---
title: "ECOL 592: NMDS"
author: "George Woolsey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
linkcolor: blue
header-includes:
  - \usepackage{caption}
  - \captionsetup[figure]{labelformat=empty}
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding){ 
    out_dir <- '/data';
    rmarkdown::render(inputFile, encoding = encoding, output_file=file.path(dirname(inputFile), out_dir, 'ECOL592_NMDS_GWoolsey.pdf')) 
  })
---

```{r setup, include=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  , fig.height = 5
  , fig.width = 7
  , eval = TRUE
  , fig.align='center'
)
# load libraries
# bread-and-butter
library(tidyverse)
library(lubridate)
library(viridis)
library(scales)
library(latex2exp)
# visualization
library(cowplot)
library(kableExtra)
# extras
library(vegan) # for NMDS
library(psych) # for correlation plot
```

# *Bombus* Example

Using data on bee (*Bombus*) biodiversity in post- bark beetle outbreak stands to explore site differences. Results were visualized using non-metric multidimensional scaling (NMDS), produced with the `metaMDS` function in package `vegan`.

Data and analysis is from [Davis et al. (2020)](https://scholar.google.com/scholar?cluster=3964994814381250489&oi=gsb&hl=en&as_sdt=0,6)

# Load Data

```{r}
data_bombus <- read.csv("DAVIS_Bombus_dataset.csv") %>% 
  # have to filter out observations with missing data
  dplyr::filter(stats::complete.cases(.)) %>% 
  # remove rows with 0's in all columns
  dplyr::mutate(r_sum = rowSums(
      dplyr::across(
        tidyselect::where(is.numeric)
        & -c(year, lat, long, SUM)
      )
    )
  ) %>% 
  dplyr::filter(r_sum>0) %>% 
  dplyr::select(-c(r_sum))
data_bombus %>% dplyr::glimpse()
#remove non-numeric columns to pass to vegan::metaMDS
data_bombus_int <- data_bombus %>% 
  dplyr::select(
    tidyselect::where(is.numeric)
    & -c(year, lat, long, SUM)
  )
data_bombus_int %>% dplyr::glimpse()
```

## Data exploration

```{r}
# correlation plot with psych
psych::pairs.panels(
  data_bombus_int %>% 
    dplyr::select(
      sample(
        x = 1:ncol(data_bombus_int)
        , size = min(8,ncol(data_bombus_int))
      )
    )
  , alpha = 0.10
  , method = "pearson" # correlation method
  , hist.col = "gray70"
  , density = TRUE # show density plots
  , ellipses = FALSE # show correlation ellipses
  , rug = FALSE
  , stars = TRUE
  , main = "sample bombas data correlations"
)
```

# NMDS

```{r}
#making the NMDS object
ordination <- vegan::metaMDS(
  data_bombus_int %>% as.matrix()
  , trymax="1000"
)
ordination
```

The NMDS results have a stress value of **`r scales::comma(ordination$stress, accuracy=0.001)`** which indicates that there are not significant differences in the clusters.

## Diagnostic plot

```{r}
#diagnostic plot
vegan::stressplot(ordination)
```

## Testing significance

```{r}
#testing for significance
#PERMANOVA
vegan::adonis2((data_bombus_int %>% as.matrix())~data_bombus$Study)	
```

## Plotting results

```{r}
# combine NMDS results with original data
data_bombus <- dplyr::bind_cols(
  ordination$points
  , data_bombus
)
# plot with color by site
plt_site <-
  ggplot(
    data = data_bombus
    , mapping = aes(x = MDS1, y = MDS2, fill = site)
  ) +
  geom_point(shape=21, alpha = 0.7, size = 4, color = "gray88") +
  scale_fill_viridis_d() +
  labs(
    title = "NMDS results colored by `site`"
  ) +
  theme_light() +
  theme(
    legend.position = "none"
  )
plt_site
# plot with color by study
plt_study <-
  ggplot(
    data = data_bombus
    , mapping = aes(x = MDS1, y = MDS2, fill = Study)
  ) +
  geom_point(shape=21, alpha = 0.7, size = 4, color = "gray88") +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    title = "NMDS results colored by `Study`"
  ) +
  theme_light() +
  theme(
    legend.position = "top"
  )
plt_study
# plot in grid
plt_study +
  facet_wrap(
    facets = dplyr::vars(Study)
    , ncol = 4
    , labeller = label_wrap_gen(width = 20, multi_line = TRUE)
  ) + 
  theme(
    legend.position = "none"
  )
# plot with color by year
plt_year <-
  ggplot(
    data = data_bombus
    , mapping = aes(x = MDS1, y = MDS2, fill = as.factor(year))
  ) +
  geom_point(shape=21, alpha = 0.7, size = 4, color = "gray88") +
  scale_fill_viridis_d(option = "cividis") +
  labs(
    title = "NMDS results colored by `year`"
  ) +
  theme_light() +
  theme(
    legend.position = "top"
  )
plt_year
# plot in grid
plt_year +
  facet_wrap(
    facets = dplyr::vars(as.factor(year))
    , ncol = 4
    , labeller = label_wrap_gen(width = 20, multi_line = TRUE)
  ) +
  theme(
    legend.position = "none"
  )

```
