---
title: 'Week 7: Ordination'
author: "Ehsan Khedive"
date: "2023-10-02"
output: word_document
---

```{r setup, include=FALSE}
#rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
Dataset <- read.csv("Data.csv")
Dataset$transsp <- sqrt((sin((pi/180*(Dataset$asp-45))/2))^2)
Dataset <- subset(Dataset,select = -asp)
Dataset$region <- as.factor(Dataset$region)
Dataset[,3:21] <- scale(Dataset[,3:21])
```

# Plotting data

```{r, fig.width=10, fig.height=10}
library(GGally)
library(ggplot2)
#ggpairs(Dataset, columns = 3:21, progress = FALSE,ggplot2::aes(colour=region))
corr_matrix <- cor(Dataset[,3:21])
library(corrplot)
testRes <- cor.mtest(corr_matrix, conf.level=0.90)
corrplot (corr_matrix, method="ellipse", p.mat=testRes$p)
```

If there is not significant correlation in a variable then potentially better to drop that variable from the PCA analysis

# PCA and Factor analysis

*Eigenvalues*, that is, the variance explained by each principal component. These are constrained to decrease monotonically from the first principal component to the last. Eigenvalues are commonly plotted on a *scree plot* to show the decreasing rate at which variance is explained by additional principal components. 

These values are called the *loadings*, which describe how much each variable contributes to a particular principal component. Large loadings (positive or negative) indicate that a particular variable strongly relates to a particular principal component. The sign of a loading indicates whether a variable and a principal component are positively or negatively correlated.

The positions of each observation in this new coordinate system of principal components are called *scores*. These are calculated as linear combinations of the original variables and the weights aij. 

```{r, fig.width=10, fig.height=10}
pcares <- princomp(Dataset[,3:21], cor = TRUE)
# # These values are called the loadings, which describe how much each variable contributes to a particular principal component. 
# Large loadings (positive or negative) indicate that a particular variable strongly relates to a particular principal component. 
# The sign of a loading indicates whether a variable and a principal component are positively or negatively correlated.
pcares$loadings
# # The positions of each observation in this new coordinate system of principal components are called scores. 
# These are calculated as linear combinations of the original variables and the weights aij.
pcares$scores
library(factoextra)
fviz_eig(pcares)
fviz_pca_ind(pcares,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
fviz_pca_var(pcares,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
# library(devtools)
# install_github("vqv/ggbiplot")
library(ggbiplot)
ggbiplot(pcares,
              obs.scale = 1,
              var.scale = 1,
              groups = Dataset$region,
              ellipse = TRUE,
              circle = TRUE,
              ellipse.prob = 0.68) + theme_bw() + scale_color_discrete(name = '') + theme(legend.direction = 'horizontal',
               legend.position = 'top')
```

# Discriminant Analysis

```{r,fig.width=10, fig.height=6}
library(MASS)
lda_model <- lda(region ~ tph+h+cov+crown+branch+shanon+regen+elev+slope+sand1+clay1+lim1+C1+K1+P1+N1+EC1+pH1+transsp, data = Dataset)

lda_model$means
lda_model$scaling

ggbiplot(lda_model,
              obs.scale = 1,
              var.scale = 1,
              groups = Dataset$region,
              ellipse = TRUE,
              circle = TRUE,
              ellipse.prob = 0.68) + theme_bw() + scale_color_discrete(name = '') + theme(legend.direction = 'horizontal',
               legend.position = 'top')+ xlab("Discriminant 1") + ylab("Discriminant 2")


predicted <- predict(lda_model, newdata = Dataset)$class
confusion_matrix <- table(predicted, Dataset$region)
confusion_matrix


```
