---
title: "ECOL 592: Spatial Correlation"
author: "George Woolsey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
linkcolor: blue
header-includes:
  - \usepackage{caption}
  - \captionsetup[figure]{labelformat=empty}
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding){ 
    out_dir <- '/';
    rmarkdown::render(inputFile, encoding = encoding, output_file=file.path(dirname(inputFile), out_dir, 'ECOL592_SpatialCorrelation_GWoolsey.pdf')) 
  })
---

```{r setup, include=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  , fig.height = 5
  , fig.width = 7
  , eval = TRUE
  , fig.align='center'
)
# load libraries
# bread-and-butter
library(tidyverse)
library(lubridate)
library(viridis)
library(scales)
library(latex2exp)
# visualization
library(cowplot)
library(kableExtra)
# extras
library(ncf) # Spatial Covariance Functions
```

# Assignment

For homework practice, please use the Bombus dataset (provided) to:

1) analyze spatial autocorrelation in abundances of one of the species (other than the one we used as an example) from one or more studies, 
2) produce a spline correlogram with a confidence interval, and 
3) interpret the correlogram to make conclusions about sample independence. How would you recommend dealing with non-spatially independent samples post-hoc? 

# Responses

## Load Data

```{r, results='hide'}
bombus <- read.csv("DAVIS_Bombus_dataset_reduced.csv", head=T) %>% 
  dplyr::rename_with(tolower)
bombus %>% dplyr::glimpse()
# which study was used for in-class presentation?
bombus[244:280,] %>% 
  dplyr::count(study)

```

## Question 1 
analyze spatial autocorrelation in abundances of one of the species (other than the one we used as an example) from one or more studies

```{r, results='hide'}
# other studies
bombus %>% 
  dplyr::count(study) %>% 
  ggplot(mapping = aes(y = study, x = n, fill = n)) +
    geom_col(width = 0.7) +
    scale_fill_viridis_c(option = "mako", direction = -1) +
    theme_light() +
    theme(legend.position = "none")
# species
bombus %>% 
  dplyr::select(
    -c(
      plot_id
      , year
      , site
      , lat
      , long
      , col_vec
    )
  ) %>% 
  tidyr::pivot_longer(
    cols = -c(study)
    , names_to = "species"
    , values_to = "count"
  ) %>% 
  dplyr::group_by(study, species) %>% 
  dplyr::summarize(n = sum(count)) %>% 
  ggplot(mapping = aes(y = species, x = n, fill = n)) +
    geom_col(width = 0.7) +
    facet_wrap(facets = vars(study)) +
    scale_fill_viridis_c(option = "mako", direction = -1) +
    scale_y_discrete(limits = rev) +
    theme_light() +
    theme(legend.position = "none")

```

\textcolor{violet}{Based on this plot of species observations by study, a correlogram will be produced only for species with observations in all studies.}

## Question 2 
produce a spline correlogram with a confidence interval

```{r}
# study list
study_list <- bombus$study %>% unique() %>% sort()
# list of species with obs from all studies
# species
species_list <- bombus %>% 
  dplyr::select(
    -c(
      plot_id
      , year
      , site
      , lat
      , long
      , col_vec
    )
  ) %>% 
  tidyr::pivot_longer(
    cols = -c(study)
    , names_to = "species"
    , values_to = "count"
  ) %>% 
  dplyr::group_by(study, species) %>% 
  dplyr::summarize(n = sum(count)) %>%
  dplyr::filter(n>0) %>% 
  dplyr::ungroup() %>% 
  dplyr::count(species) %>% 
  dplyr::filter(n == length(unique(bombus$study))) %>% 
  dplyr::pull(species)
# function to plot correlogram
plt_fn <- function(study_num = 1, species_num = 1) {
  # filter data
  dta_temp <- bombus %>% 
    dplyr::filter(study == study_list[study_num])
  # spline.correlog
  correlog_temp <- ncf::spline.correlog(
    x=dta_temp$long
    , y=dta_temp$lat
    , z=dta_temp[[species_list[species_num]]]
    , latlon=T
    , resamp=200
    , quiet = T
  )
  # set up data frame and plot
  plt <- data.frame(
      x = correlog_temp$real$predicted$x %>% c()
      , y = correlog_temp$real$predicted$y %>% c()
      , l_ci = correlog_temp$boot$boot.summary$predicted$y["0.025",] %>% c()
      , u_ci = correlog_temp$boot$boot.summary$predicted$y["0.975",] %>% c()
    ) %>% 
    dplyr::filter(x < correlog_temp$max.distance) %>% 
    ggplot(
      mapping = aes(x = x, y = y)
    ) + 
    geom_hline(yintercept = 0, color = "black") +
    geom_ribbon(aes(ymin = l_ci, ymax = u_ci), fill = "grey70", alpha = 0.8) +
    geom_line(color = "navy") + 
    labs(
      x="distance"
      , y = "correlation"
      , title = paste0("study: ",study_list[study_num])
      , subtitle = paste0("species: ",species_list[species_num])
    ) +
    theme_light() + 
    theme(
      plot.title = element_text(size = 9)
      , plot.subtitle = element_text(size = 8)
      , axis.title = element_text(size = 6)
      , axis.text = element_text(size = 6)
    )
  # return
  return(plt)
}
# plt_list <- c("Exp","Sph","Gau","Exc","Mat","Log","Pow","Spl") %>% 
plt_list <- 1:length(study_list) %>% 
  purrr::map(plt_fn)
# combine
cowplot::plot_grid(
  plotlist = plt_list 
  , ncol = 3
)
```

\textcolor{violet}{The figure above shows a spline correlogram for all studies with observations of the species: }`r species_list[species_num]`. 

## Question 3
interpret the correlogram to make conclusions about sample independence. How would you recommend dealing with non-spatially independent samples post-hoc? 

\textcolor{violet}{Interpretation of these plots should be based on the goodness test.}